Chapter 06: National Education Trends Case Study
========================================================

First let's go ahead and download our data from GitHub:

```{r}
# This will let us download the file from a remote URL

download.file("https://raw.githubusercontent.com/faulconbridge/appliedStats/master/LaTeX/part03/data/correlationCaseStudy.csv","census.csv","wget",extra="--no-check-certificate")

# And now we will read the data into R and store it
# in a data frame called 'data'

data <- read.csv("census.csv",header=TRUE,sep=",")
```

Now, we have two columns that we're interested in: highSchoolorHigher and perCapitaIncome. We would like to see if there is a relation between a state's per capita income and the proportion of its residents to have completed high school or higher. So let's start by constructing a scatterplot of the two variables:

```{r}
# Notice, when we reference a column we use the syntax
# dataset$column. The first half tells R which data set
# we are referencing, the dollar sign indicates that we
# want to reference a specific column, and everything
# after that is the column name itself.

plot(data$HighSchoolOrHigher,data$perCapitaIncome,
     xlab="Proportion with HS diploma or higher",
     ylab="Per capita income (dollars)",
     main="Scatterplot of Education by Per Capita Income",
     pch = 21, bg = "dodgerblue2", cex.axis = 0.8,
     cex.lab = 0.9, cex.main = 0.9)

# We can also plot a line of best fit:

fit <- lm(data$perCapitaIncome~data$HighSchoolOrHigher)
abline(fit, col = "dodgerblue4")
```

As we can see, there appears to be a positive linear relationship between per capita income and the proportion of a state's residents having a high school diploma or higher. Our next step is then to quantify the strength of this relationship. To do this, we will perform a bivariate correlation

```{r}
# The syntax for a correlation test in R is easy:
# it is cor.test(var1, var2) where var1 and var2
# are the two variables that you're interested in.

cor.test(data$perCapitaIncome,data$HighSchoolOrHigher)
```

As we can see, there are several noteworthy items presented in this table. Firstly, we have a <em>t</em>-statistic of 2.78. This is above the 1.96 threshold that we set, indicating that our correlation is probably going to be significant. From there, we can look at our <em>p</em>-value (0.008) and see that it is below the 0.05 threshold, indicating that we do indeed have a significant correlation.

If it weren't obvious from the scatterplot above, this is a positive correlation with a Pearson's r=0.37, meaning that there exists a fair positive relationship between our two variables. I.e., when one is larger, the other will also tend to be larger.